---
title: "dpkg: Data Packages for R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
library(magrittr)
```

An R package to read, write, and edit [Data Package](https://specs.frictionlessdata.io/data-package/) data and metadata. Unlike other existing R packages [dpmr](https://github.com/christophergandrud/dpmr) and [datapkg](https://github.com/ropenscilabs/datapkg), dpkg can be used to build and document Data Packages entirely within R. Please note that this is a work in progress and function naming and functionality may drift based on feedback from the community.

This package is not on CRAN. To install in R, use [devtools](https://github.com/hadley/devtools):

```
devtools::install_github("ezwelty/dpkg")
```

#### Quick introduction

To build a data package, assemble the data and add metadata to the various elements:

```{r}
data <- data.frame(
  id = 1L %>% set_field(title = "Identifier"),
  value = 1.1,
  added = Sys.Date()
)
# Data Resource (list of Fields)
dr <- data %>%
  set_resource(
    name = "data",
    path = "data/data.csv"
  )
# Data Package (list of Resources)
dp <- list(dr) %>%
  set_package(
    name = "data-package"
  )
```

You can preview the package metadata:

```{r}
get_package(dp) %>% str()
```

Write the package to file:

```{r}
dir <- tempdir()
write_package(dp, path = dir)
```

And read the package back in:

```{r}
read_package(dir)
```

## Build a package

In `dpkg`, the contents of a data *package* is stored as a list of one or more data *resources* (each a list) of one or more *fields* (each typically an atomic vector). For example:

```{r}
dp <- list(
  dr = data.frame(
    id = 1L,
    value = 1.1,
    added = Sys.Date()
  )
)
```

Package, resource, and field ("data objects") metadata can be set or updated using the `set_*` functions (`set_package`, `set_resource`, `set_field`), which come in a `<-` flavor:

```{r}
set_field(dp$dr$id) <- field(title = "Unique identifier", constraints = constraints(unique = TRUE))
```

and a pipe-friendly flavor:

```{r}
dp$dr$id %<>% set_field(title = "Identifier", constraints = NULL)
```

As seen above with the use of `field` and `constraints`, a suite of helper functions are available to assist in the building of metadata:

  * Data objects: `package`, `resource`, `field`
  * Meta objects:  `schema`, `foreignKey`, `constraints`, `license`, `source`, `contributor`

## Preview a package

To preview a package, metadata can be retrieved from data objects using the `get_*` functions (`get_package`, `get_resource`, `get_field`). Missing properties are filled with their default values:

  * Fields
      * `name`: The name of the object in a list (resource).
      * `type`: The type corresponding to the object class.
          * `character` -> `"string"`
          * `numeric` -> `"number"`
          * `integer` -> `"integer"`
          * `logical` -> `"boolean"`
          * `Date` -> `"date"`
          * `POSIXt` -> `"datetime"`
          * otherwise -> `"string"`
      * `format`: The default format for that type.
          * `date` -> `"%Y-%m-%d"`
          * `datetime` -> `"%Y-%m-%dT%H-%M-%SZ"`
          * `unit`: Units set by [units](https://github.com/edzer/units) deparsed to product power form.
  * Resources
      * `name`: The name of the object in a list (package).
      * `schema$fields`: Field metadata from the elements of the object.
  * Packages
      * `resources`: Resource metadata from the elements of the object.

```{r}
get_field(dp$dr$id) %>% str()
get_resource(dp$dr) %>% str()
get_package(dp) %>% str()
```

## Write a package

`write_package` writes package data and metadata to disk using the following rules for each resource:

  * `format`: If missing, checks `path` file extension and `mediatype`. Only "csv" ("text/csv") and "json" ("application/json") are supported.
  * `path`: If not set, the data is saved in the metadata (`datapackage.json`) as either an inline JSON object (`format:` "json" or missing) or a CSV string (`format:` "csv"). For writing, `path` must be a single, local, relative path.

```{r, echo=FALSE}
tmpdir <- file.path(tempdir(), "example")
```

Resource as an inline JSON object:

```{r}
set_resource(dp$dr) <- package(format = "json", path = NULL)
get_resource(dp$dr)$data
write_package(dp, path = tmpdir)
list.files(tmpdir)
```
```{r, echo=FALSE}
unlink(tmpdir, recursive = TRUE)
```

Resource as an inline CSV string:

```{r}
set_resource(dp$dr) <- package(format = "csv", path = NULL)
get_resource(dp$dr)$data
write_package(dp, path = tmpdir)
list.files(tmpdir)
```
```{r, echo=FALSE}
unlink(tmpdir, recursive = TRUE)
```

Resource as a JSON file:

```{r}
set_resource(dp$dr) <- package(format = "json", path = "data/data.json")
get_resource(dp$dr)$data
write_package(dp, path = tmpdir)
list.files(tmpdir, recursive = TRUE)
```
```{r, echo=FALSE}
unlink(tmpdir, recursive = TRUE)
```

Resource as a CSV file:

```{r}
set_resource(dp$dr) <- package(format = "csv", path = "data/data.csv")
get_resource(dp$dr)$data
write_package(dp, path = tmpdir)
list.files(tmpdir, recursive = TRUE)
```
```{r, echo=FALSE}
unlink(tmpdir, recursive = TRUE)
```

## Read a package

`read_package` reads package data and metadata into the same structure described above, but unlike `write_package`, it supports both local and remote paths. The `resources` argument can be used to read a subset of the package's resources (or all if `NULL`, the default).

```{r, message=FALSE, warning=FALSE}
dp <- read_package(
  "https://raw.githubusercontent.com/columbia-glacier/optical-surveys-1985/master",
  resources = c("station", "velocity")
)
get_package(dp) %>% str()
dp$station
head(dp$velocity)
```

`read_package_github` accepts a shorthand GitHub repository address.

```{r, message=FALSE, warning=FALSE}
dp <- read_package_github("columbia-glacier/optical-surveys-1985", "station")
```

## TODO

### Fields

Only types `string`, `number`, `integer`, `boolean`, `date`, and `datetime` are implemented (see [table-schema/field-descriptors](https://specs.frictionlessdata.io/table-schema/#field-descriptors)). Add support for the remaining types:

  - [ ] `type =` [object](https://specs.frictionlessdata.io/table-schema/#object)
  - [ ] `type =` [array](https://specs.frictionlessdata.io/table-schema/#array)
  - [ ] `type =` [time](https://specs.frictionlessdata.io/table-schema/#time) (via package [hms](https://github.com/tidyverse/hms))
  - [ ] `type =` [year](https://specs.frictionlessdata.io/table-schema/#year) (already supported via `type = date` and `format = "%Y"`)
  - [ ] `type =` [yearmonth](https://specs.frictionlessdata.io/table-schema/#yearmonth) (already supported via `type = date` and `format = "%Y-%m"`)
  - [ ] `type =` [duration](https://specs.frictionlessdata.io/table-schema/#duration) (already supported via `type = numeric` and `unit`)
  - [ ] `type =` [geopoint](https://specs.frictionlessdata.io/table-schema/#geopoint)
  - [ ] `type =` [geojson](https://specs.frictionlessdata.io/table-schema/#geojson)

Additionally:

  - [ ] Validate field values against [`constraints`](https://specs.frictionlessdata.io/table-schema/#constraints) property
  - [ ] For `type =` [string](https://specs.frictionlessdata.io/table-schema/#string), validate values against `format` property
  - [ ] For `type =` [number](https://specs.frictionlessdata.io/table-schema/#number), support `currency` property - *What does this property contain?*

### Resources & Packages

  - [ ] Validate packages against the default or custom [`profile`](https://specs.frictionlessdata.io/profiles/) - *The published schemas don't match the latest specifications?!*
  - [ ] Read/write resources with a `path` like "data/data.csv.gz" to/from compressed files
  - [ ] Write resource schemas with a `path` to a JSON file
  - [ ] Read/write GeoJSON and TopoJSON to/from spatial objects
